---
title: "HPI Data: from excel to ggplot/plotly via dplyr"
output: html_notebook
---

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(tidyquant)
library(highcharter)
library(readxl)
```

Here's a data import and wrangling exercise. Our purpose is to take an excel file of the housing price index (hpi) for the 50 states since 1975 and get into a format that can be used to visualize the data using ggplot, highcharter and leaflet.

For `ggplot` we'll need to get this excel data into a data.frame or tibble with an appropriate date column. 

For `highcharter`, we'll need an `xts` object. 

```{r, message = FALSE, warning = FALSE}

# Load in data frame of housing price index for all 50 states
# Plus the national average
#states_hpi <- 
 # Quandl("FMAC/HPI", order = 'asc')

# Use the Import Dataset button to show how to do some 
# pre-sorting. After import, we'll have a data.frame object
# that still needs work - we'll wrangle it into shape.
states_hpi <- read_excel("states_hpi.xls", 
    col_types = c("text", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "skip", "skip"), na = "NA", skip = 5)

states_wrangled_highcharter <-
  states_hpi %>%
  select(-53) %>% 
  # remove confusing metadata
  slice(-508:-523) %>% 
  # add nice date column
  mutate(Date = seq(mdy('01/31/1975'),mdy('03/31/2017'), by = 'months')) %>%
  # remove the month column, move date column to be first
  select(Date, everything(), -Month) %>% 
  as_xts(date_col = Date)

head(states_wrangled_highcharter)
```

```{r}
# We don't need to, but if we needed to go back to a tibble/data frame? 
states_back_to_tibble <- 
  states_wrangled_highcharter %>% 
  as_tibble(preserve_row_names = TRUE) %>%
  mutate(date = ymd(row.names)) %>% 
  select(-row.names) %>% 
  select(date, everything())
```

```{r}

highchart(type = "stock") %>%
  hc_title(text = "California and Georgia HPI") %>%
  hc_add_series(states_wrangled_highcharter$GA, name = "Georgia HPI", color = 'blue') %>%
  hc_add_series(states_wrangled_highcharter$CA, name = "California HPI", color = 'green') %>%
  hc_navigator(enabled = FALSE) %>% 
  hc_scrollbar(enabled = FALSE)

```



