---
title: "R Notebook"
output: html_notebook
---

```{r setup, message = FALSE, warning = FALSE}
library(tigris)
library(tidyverse)
library(tidyquant)
library(leaflet)
library(highcharter)
library(readxl)
```

```{r, warning = FALSE}
states <- states(cb = TRUE, class = "sf")
```

Here's a data import and wrangling exercise. Our purpose is to take an excel file of the housing price index (hpi) for the 50 states since 1975 and get into a format that can be used to visualize the data using ggplot, highcharter and leaflet.

For `ggplot` we'll need to get this excel data into a data.frame or tibble with an appropriate date column. 

For `highcharter`, we'll need an `xts` object. 

`leaflet`: we'll see.

```{r, message = FALSE, warning = FALSE}

# Load in data frame of housing price index for all 50 states
# Plus the national average
#states_hpi <- 
 # Quandl("FMAC/HPI", order = 'asc')

# Use the Import Dataset button to show how to do some 
# pre-sorting. After import, we'll have a data.frame object
# that still needs work - we'll wrangle it into shape.
states_hpi <- read_excel("states_hpi.xls", 
    col_types = c("text", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "skip", "skip"), na = "NA", skip = 5)

# Take a look at the states_hpi object. 
# The data column is called "month", it's terribly, unusably formatted.
# We'll use `mutate` to create a 'date' column.
# There are 15 rows of metadata at the bottom. Need to remove those with `slice`

states_hpi_wrangled_leaflet <- 
  states_hpi %>% 
  # remove confusing metadata
  slice(-508:-523) %>% 
  # add nice date column
  mutate(Date = seq(mdy('01/31/1975'),mdy('03/31/2017'), by = 'months')) %>%
  # remove the month column, move date column to be first
  select(Date, everything(), -Month) %>% 
  # Filter to just the last year (March to March)
  filter(Date == ymd("2016-03-31") | Date == ymd("2017-03-31")) %>% 
  # Change to long format
  gather(state, value, -Date) %>% 
  # Groupd by state
  group_by(state) %>%
  # Add annual price appreciation value with mutate and lag function
  mutate(hpa = (value - lag(value))/ lag(value)) %>%
  # Round that new value
  mutate(hpa = round(hpa, digits = 3) * 100)  %>% 
  # good practice to get rid of any NAs, unless those are important to identify
  na.omit() %>% 
  # Get rid all except state and HPA. We don't care about the rest.
  select(state, hpa) %>% 
  # Rename the column? Why in the world choose STUSPS? 
  # So we can add it to our shapefile/map
  rename(STUSPS = state)

```

```{r}
states_wrangled_ggplot <- 
  states_hpi %>% 
  # remove confusing metadata
  slice(-508:-523) %>% 
  # add nice date column
  mutate(Date = seq(mdy('01/31/1975'), mdy('03/31/2017'), by = 'months')) %>%
  # remove the month column, move date column to be first
  select(Date, everything(), -Month) %>% 
  ##gather from wide to long format
  gather(state, hpi, -Date) %>%
  ##round hpi to four digits
  mutate(hpi = round(hpi, digits = 4)) %>%
  ##create new column: hpa, hpa12, hpa3 etc.
  mutate(
         hpa = ((1+ (hpi-lag(hpi))/hpi)^12) - 1, 
         hpa12 = ((1+ (hpi-lag(hpi, 12))/hpi)^1) - 1, 
         hpa3 = ((1+ (hpi-lag(hpi, 3))/hpi)^4) - 1, 
         hpi12 = lag(hpi, 12), 
         hpi12min = rollapply(hpi, 12, min, fill = NA, na.rm = FALSE, align = 'right'),
         hpi12max = rollapply(hpi, 12, max, fill = NA, na.rm = FALSE, align = 'right'))
```

```{r, message = FALSE, warning = FALSE}
## Work with data from 1999-2017
states_post_1990 <- 
  states_wrangled_ggplot %>%
  filter(Date >= "1990-01-01" & Date <= "2010-01-01")

# Drop the states_post_1990 object into the a ggplot

hpa_ggplot <- 
  ggplot(states_post_1990, aes(Date)) +
  geom_line(aes(y = hpa), colour = "blue") +
  #geom_line(aes(y=hpi12max), colour = "red") +
  #geom_point(data = subset(states_post_1990, Date == "2008-05-31"), 
   #          aes(x = Date, y = hpa), color = "green", alpha = 0.7) +
  ylab("Housing Price Index") +
  #facet_wrap(~state, ncol = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  theme_minimal()

hpa_ggplot
ggplotly(hpa_ggplot)
```


```{r}
# Now we want to merge by a common column name. 
states_hpa_leaflet <- merge(states, states_hpi_wrangled, by = "STUSPS", all.x = TRUE)
# There's now a column in our map called hpa. Why is there an NA? 
head(states_hpa_leaflet$hpa)
head(states_hpa_leaflet$NAME)
# There's an NA becuase American Somoa is not a state and we only data about states. That NA is not going
#hurt us. Instead, it won't get a nice color on the map. Why else is that helpful? If missed a state, it 
# would be grey.
```


```{r, fig.height = 5, fig.width = 10, message = FALSE, warning = FALSE}
# Build states map
statesPal<-colorNumeric(
  palette = "GnBu",
  domain = states_hpa_leaflet$hpa)

statesPopup <- paste0( 
                    states_hpa_leaflet$NAME,
                    "<br>Annual House Price Percent Change: ", 
                     states_hpa_leaflet$hpa, "%")

leaf_states <- 
  leaflet(states_hpa_leaflet) %>% 
  addProviderTiles("CartoDB.Positron") %>%   
  setView(-95, 40, zoom = 4) %>% 
  addPolygons(stroke = TRUE, color = "black", weight = .4, opacity = 1.0, 
              smoothFactor = 0.5, fill = TRUE, fillColor = ~statesPal(hpa), 
              fillOpacity = .8, layerId = ~STUSPS, popup = statesPopup)

leaf_states
```


```{r}
states_hpi_highcharter <-
  states_hpi %>%
  select(-53) %>% 
  # remove confusing metadata
  slice(-508:-523) %>% 
  # add nice date column
  mutate(Date = seq(mdy('01/31/1975'),mdy('03/31/2017'), by = 'months')) %>%
  # remove the month column, move date column to be first
  select(Date, everything(), -Month) 

# We need to convert to xts before calling highcharter. 
states_hpi_xts <- 
  states_hpi_highcharter %>% 
  as_xts(date_col = Date)

# We don't need to, but if we needed to go back to a tibble/data frame? 
states_back_to_tibble <- 
  states_hpi_xts %>% 
  as_tibble(preserve_row_names = TRUE) %>%
  mutate(date = ymd(row.names)) %>% 
  select(-row.names) %>% 
  select(date, everything())

highchart(type = "stock") %>%
  hc_title(text = "California and Georgia HPI") %>%
  hc_add_series(states_hpi_xts$GA, name = "Georgia HPI", color = 'blue') %>%
  hc_add_series(states_hpi_xts$CA, name = "California HPI", color = 'green') %>%
    # I don't like the look of the navigator/scrollbar, but you might. 
    # Change these to enabled = TRUE and check out the results.
  hc_navigator(enabled = FALSE) %>% 
  hc_scrollbar(enabled = FALSE)

```

```{r}
save(leaf_states, file = "leaflet_states.RDat")
```

